\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-16 docs/parco13/abstract.tex
Move my literature too and case change
Comments

\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{color}
\usepackage{xspace}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement, especially when working with
unstructured meshes \comment{This sounds we are now going to talk
  about unstructured meshes.}.  Of course, it has been realized early
that it helps to aggregate degrees of freedom (DOF) at the element
level, as it has been done with high-order spectral element
\cite{TufoFischer99}, finite volume
\cite{paramesh, chombo, samrai}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.
\comment{Are you looking for references to parallel AMR codes?  Or more general AMR codes based on the FVM? }

@@

To enable hardware acceleration for parallel dynamic AMR, we like to
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In this work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989, be-le:1991,
ColellaGravesKeenEtAl07} except that the
patches are not overlapping, which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The Clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
\comment{I am not sure about the Python dependency;  the latest version of Clawpack relies on
Python to generate an input file, but the code itself is F77 (or maybe some F90). }
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (approx 100 bytes per leaf
\comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}) is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).
\end{enumerate}

@@

\forestclaw uses iterators over all leafs, optionally restricted to a given
level.  Random access is possible and used when executing $O(1)$ time neighbor
lookups.  Looping over the patches in the order prescribed by the forest \comment{tree?} and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse.
% since the ghost patches are likely closeby with respect to the $z$-order.

@@

\todo{What is BearClaw (tree-based / S.\ Mitran)?}

@@

%%%%%%%%%%% The bibliography starts:
\bibliographystyle{unsrt}
\bibliography{ccgo,../Biblio/amr}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{color}
\usepackage{url}
\usepackage{xspace}

@@

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it has been realized early
that it helps to aggregate degrees of freedom (DOF) at the element
level, as it has been done with high-order spectral element
\cite{TufoFischer99}, finite volume
\cite{paramesh, chombo, samrai}, or discontinuous Galerkin
\cite{BursteddeGhattasGurnisEtAl10} methods.
\comment{Are you looking for references to parallel AMR codes?  Or more general AMR codes based on the FVM?}
\todo{No, I was looking for unstructured FV references.  Would you have any,
and can we add the block-AMR references later?}

@@

To enable hardware acceleration for parallel dynamic AMR, we like to
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989, be-le:1991,
ColellaGravesKeenEtAl07,Uintah} except that the
patches are not overlapping, which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The Clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% Clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (24 bytes per \pforest leaf plus 92 bytes per
forestclaw patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I hav added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).
\end{enumerate}

@@

\forestclaw uses iterators over all leafs, optionally restricted to a given
level.  Random access is possible and used when executing $O(1)$ time neighbor
lookups.  Looping over the patches in the order prescribed by the forest
% \comment{tree?}
% CB: I think forest is fine here.
and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse
\cite{BursteddeBurtscherGhattasEtAl09}.
% since the ghost patches are likely closeby with respect to the $z$-order.

@@

% \todo{What is BearClaw (tree-based / S.\ Mitran)?}

@@

%%%%%%%%%%% The bibliography starts:
\bibliographystyle{unsrt}
\bibliography{../biblio/ccgo,../biblio/amr}
